---
title: "assignment"
author: "swimbait"
date: "May 1, 2020"
output: html_document
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading and preprocessing the data
```{r echo=TRUE}
activity<-read.csv("activity.csv")
head(activity)
tail(activity)
```
## What is mean total number of steps taken per day?
```{r echo=TRUE}
stepSumByDay<-tapply(activity$steps,activity$date,FUN=function(x) sum(x, na.rm=T))
meanStepsPerDay<-round(mean(stepSumByDay),0)
medianStepsPerDay<-round(median(stepSumByDay),0)
hist(stepSumByDay)
```

The mean steps per day is `r round(meanStepsPerDay,0)` and the median steps per day is `r round(medianStepsPerDay,0)`.

## What is the average daily activity pattern?
```{r echo=TRUE}
meanStepsByInterval<-tapply(activity$steps,activity$interval,FUN=function(x) mean(x,na.rm=T))
plot(names(meanStepsByInterval), meanStepsByInterval,type="l", xlab="Within-Day Interval")
maxInterval<-names(meanStepsByInterval)[which(meanStepsByInterval==max(meanStepsByInterval))]
```

The interval which contains the most steps per day is `r maxInterval`

## Imputing missing values
```{r echo=TRUE}
missingSteps<-which(is.na(activity$steps))
meanStepsByIntervalFrame<-data.frame(cbind(interval=names(meanStepsByInterval),meanSteps=round(meanStepsByInterval,0)))
missingStepsActivity<-activity[missingSteps,]
activityImputed<-merge(activity, meanStepsByIntervalFrame,by="interval",all.y=F)
activityImputed<-activityImputed[order(activityImputed$date,activityImputed$interval),]
activityImputed$imputedSteps<-ifelse(is.na(activityImputed$steps),as.numeric(as.vector(activityImputed$meanSteps)),activityImputed$steps)
stepSumByDay2<-tapply(activityImputed$imputedSteps,activityImputed$date,FUN=function(x) sum(x, na.rm=T))
meanStepsPerDay2<-round(mean(stepSumByDay2),0)
medianStepsPerDay2<-round(median(stepSumByDay2),0)
hist(stepSumByDay2)

```

Missing step data are imputed by with the mean value for that within-day interval across all days in the data.  

The total amount of within date intervals with missing values on steps is `r sum(is.na(activity$steps))`.
The mean steps per day is `r round(meanStepsPerDay2,0)` and the median steps per day is `r round(medianStepsPerDay2,0)`.

The imputed missing values has a large impact on the mean but not as strong of an impact on the median.

## Are there differences in activity patterns between weekdays and weekends?
```{r echo=TRUE}
activityImputed$dayOfWeek<-weekdays(as.Date(activityImputed$date))
activityImputed$dayType<-as.factor(ifelse(activityImputed$dayOfWeek %in% c("Sunday","Saturday"),"Weekend","Weekday"))
dayType<-activityImputed$dayType=="Weekend"
meanStepsByIntervalWeekend<-tapply(activityImputed$steps[dayType],activity$interval[dayType],FUN=function(x) mean(x,na.rm=T))
meanStepsByIntervalWeekday<-tapply(activityImputed$steps[!dayType],activity$interval[!dayType],FUN=function(x) mean(x,na.rm=T))
plot(names(meanStepsByIntervalWeekend), meanStepsByIntervalWeekend,type="l", xlab="Within-Day Interval",lwd=1.5)
lines(names(meanStepsByIntervalWeekday), meanStepsByIntervalWeekday,col="red")
```

Weekend average steps by interval is in black and weekday is red
